---
title: "Lab 07 — Model Selection and Validation"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(broom)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab extends your linear modeling work from **fitting and interpreting regression models** to **comparing models**.

In Part I, you will evaluate model fit using:

- **sum of squares**
- **R-squared**
- **MSE**
- **RMSE**
- and **residual diagnostics**

You will work with the **US-AMS simplified AmeriFlux dataset** again and choose your own response and predictor variables.

Part II will later extend this work to **train-test validation splits**.

---

# Learning goals

By the end of this lab, you should be able to:

- Fit and compare multiple linear regression models
- Use `tidy()`, `glance()`, and `augment()` to summarize models
- Compute **SST**, **SSE**, and **R-squared**
- Compute **MSE** and **RMSE**
- Compare models using both statistical summaries and plots
- Explain why model selection requires more than one metric
- Produce a fully reproducible R Markdown analysis

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=TRUE}
# GOAL: Read the dataset and keep 2023 observations.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns if needed

```{r derive_time, echo=TRUE, eval=TRUE}
# GOAL: Create date / month columns if they do not already exist.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE)
  )
```

---

# Part I — Model selection and assessment

## 1) Choose a response variable and candidate predictors

Pick:
- **one numeric response variable**
- **two candidate predictor variables**

Possible examples:
- response: `le`, `fc`, `gpp`
- predictors: `ta`, `swc`, `vpd`, `rg`

```{r choose_vars, echo=TRUE, eval=TRUE}
response_var <- "gpp"
pred1 <- "swc"
pred2 <- "ta"

# CHECK: do these variables exist?
c(response_var, pred1, pred2) %in% names(df)
```

> Which variables did you choose, and why do you think they may be related?

I chose `gpp` as my response variable and `swc` and `ta` as my predictor variables. I chose `ta` and `swc` as predictor values for `gpp` because plant growth and productivity depend upon the the amount of water plants receive or have access to, as well as the temperature of the environment they are in. If it is too hot or too cold, plants might not be able to survive. Furthermore, they both change seasonally, like `gpp`, meaning they might be variables that impact `gpp`. For example, `ta`, or air temperature, is usually warmer in the spring and summer, when `gpp` tends to be  highest, than in the fall and winter, when `gpp` tends to be lowest.

---

## 2) Create a clean analysis dataset

Keep only rows with non-missing values for your chosen variables.

```{r make_analysis_df, echo=TRUE, eval=TRUE}
# GOAL: Create an analysis data frame using only complete cases.

df_mod <- df %>%
  select("gpp", "swc", "ta") %>%
  drop_na()

glimpse(df_mod)
```

---

## 3) Fit three models

Fit:
- **Model 1:** response ~ pred1
- **Model 2:** response ~ pred2
- **Model 3:** response ~ pred1 + pred2

```{r fit_models, echo=TRUE, eval=TRUE}
m1 <- lm(gpp ~ swc, data = df_mod)
m2 <- lm(gpp ~ ta, data = df_mod)
m3 <- lm(gpp ~ swc + ta , data = df_mod)
```

---

## 4) View coefficients with `tidy()`

```{r tidy_models, echo=TRUE, eval=TRUE}
tidy(m1)
tidy(m2)
tidy(m3)
```

> Which coefficients are positive? Which are negative? Do they make scientific sense?

The coefficient of `ta` is positive, whereas the coefficient of `swc` is negative. The coefficient for `swc` does not make scientific sense, but the coefficient for `ta` does make sense. 

---

## 5) Visualize at least two one-predictor models

Make scatterplots with fitted linear trends for Model 1 and Model 2.

```{r plot_m1, echo=TRUE, eval=TRUE}
ggplot(df_mod, aes(x = .data[[pred1]], y = .data[[response_var]])) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic(base_size = 18) +
  labs(
    x = pred1,
    y = response_var,
    title = "Model 1"
  )
```

```{r plot_m2, echo=TRUE, eval=TRUE}
ggplot(df_mod, aes(x = .data[[pred2]], y = .data[[response_var]])) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic(base_size = 18) +
  labs(
    x = pred2,
    y = response_var,
    title = "Model 2"
  )
```

> Visually, which model seems stronger? Why?

Model 2 seems visually stronger. This is because the linear relationship is clearer and appears to be stronger in Model 2 than in Model 1. Model 2 also has more structure to the values scattered around the line than Model 1. 

---

## 6) Get fitted values and residuals with `augment()`

```{r augment_models, echo=TRUE, eval=TRUE}
aug_m1 <- augment(m1, data = df_mod)
aug_m2 <- augment(m2, data = df_mod)
aug_m3 <- augment(m3, data = df_mod)

aug_m1 %>% select(.fitted, .resid) %>% head()
```

---

## 7) Compute sum of squares manually for Model 1

Recall:

- SST = total sum of squares
- SSE = sum of squared errors
- SSR = regression sum of squares

And:

$SST = SSR + SSE$

```{r ss_model1, echo=TRUE, eval=TRUE}
y <- df_mod[[response_var]]
ybar <- mean(y)

sst <- sum((y - ybar)^2)
sse_m1 <- sum(aug_m1$.resid^2)
ssr_m1 <- sst - sse_m1

tibble(
  sst = sst,
  sse_m1 = sse_m1,
  ssr_m1 = ssr_m1
)
```

> In words, what do SST and SSE represent?

`SST` represents the total variation in the response variable `gpp`. `SSE` represents the variation in the response variable `ggp` that is not explained by the model. 

---

## 8) Compute R-squared manually for Model 1

Use:

$R^2 = 1 - SSE / SST$

```{r r2_manual, echo=TRUE, eval=TRUE}
r2_m1_manual <- 1 - sse_m1 / sst

r2_m1_manual
```

## 9) Compare to `glance()`

```{r glance_model1, echo=TRUE, eval=TRUE}
glance(m1)
```

> Does your manual R-squared match the value from `glance()`?

Yes my manual R-squared (0.1199387) matches the value from `glance()` (0.1199387).

---

## 10) Compute MSE and RMSE for Model 1

Use:

- $MSE = mean(e_i^2)$
- $RMSE = sqrt(MSE)$

```{r mse_rmse_m1, echo=TRUE, eval=TRUE}
mse_m1 <- mean (aug_m1$.resid^2)
rmse_m1 <- sqrt(mse_m1)

tibble(
  mse_m1 = mse_m1,
  rmse_m1 = rmse_m1
)
```

> What does RMSE mean in the units of your response variable?

The expected error of the models predictions is 9.723972 µmolCO2 m-2 s-1 away from the actual `gpp` values.

---

## 11) Compare all three models using `glance()`

```{r compare_glance, echo=TRUE, eval=TRUE}
bind_rows(
  glance(m1) %>% mutate(model = "m1"),
  glance(m2) %>% mutate(model = "m2"),
  glance(m3) %>% mutate(model = "m3")
) %>%
  select(model, r.squared, adj.r.squared, sigma, AIC, BIC)
```

> Which model has the highest R-squared? Is that enough to call it the best model?

Model `m3` has the highest r-squared. That is not enough to call it the best model. 

---

## 12) Compare all three models using RMSE

```{r compare_rmse, echo=TRUE, eval=TRUE}
rmse_tbl <- bind_rows(
  augment(m1, data = df_mod) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m1"),
  augment(m2, data = df_mod) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m2"),
  augment(m3, data = df_mod) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m3")
)

rmse_tbl
```

> Which model has the smallest RMSE? Does that agree with the R-squared ranking?

Model `m3` has the smallest RMSE. That does agree with the R-squared ranking. 

---

## 13) Compare residual plots

Make residual vs fitted plots for at least two models.

```{r resid_plot_m1, echo=TRUE, eval=TRUE}
ggplot(aug_m1, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic(base_size = 18) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals: Model 1"
  )
```

```{r resid_plot_m3, echo=TRUE, eval=TRUE}
ggplot(aug_m3, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic(base_size = 18) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals: Model 3"
  )
```

> Do the residual plots suggest any problems (curvature, changing variance, outliers)? Which model looks more appropriate?

The residual plots do suggest problems. Model 1 looks more appropriate because its residuals are more randomly scattered around zero, suggesting no strong curvature. This means that for Model 1 the assumption of linearity is met. However, the residual plot for Model 3 shows a strong curvature. Furthermore, Model 3 displays a fan-shaped, suggesting strong heteroscedasticity. This means that for Model 3 the assumption of lineararity is violated. 

---

## 14) Make a model selection decision

Choose the model you think is best supported by the evidence.

Fill in:

> My preferred model is `Model 3`

> I based this decision on:
> - `Model 3 has the highest r-squared value`
> - `Model 3 has the lowest RMSE value`
> - `Model 3 has residuals that are generally centered around zero`

**Prompt (5–7 sentences):** Explain why you chose this model. Use at least two quantitative criteria (such as R-squared, RMSE, or sigma) and one visual criterion (such as the residual plot or scatterplot). Also explain why a different model was less convincing.

> I chose Model 3 because it had the highest R-squared value, meaning the model explained more of the variability in `gpp` than the other models. I also chose Model 3 because it had the lowest RMSE value, meaning its prediction error is smaller than the other models. This suggests that the Model 3's predicted values where closer to the actual values than the other models predicted values were. Finally, I chose Model 3 because, despite the issues in linearity in its residual plot, the plot revealed that the residuals where still generally centered around zero. Model 3 had smaller residuals than the other models, indicating that the predicted values where close to the actual values. Model 1 was less convincing because it had a low R-squared value and a high RMSE value. Furthermore, Model 1 also had higher residuals than Model 3, despite the assumption of linearity being met.  

---

# Part II — Placeholder: train-test validation

In the next lecture and lab section, we will extend model selection into **validation**.

We will ask:

- Does a model that fits the training data well also predict new data well?
- How do train-test splits change model comparison?
- When can a more complex model overfit?

> *Instructor will add this section later. Do not delete this header.*

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
